Summary of tumor cell classification with `AUCell` for `r params$library_id`

### 2024-07-26 analyses/cell-type-ewings/template_notebooks/auc-singler-workflow/01-auc-results.Rmd

# test if _any_ marker genes have expression
if(sum(markers_df$gene_expression) == 0){
  has_marker_gene_exp <- FALSE
  message("No marker gene expression detected, so all plots of marker genes will be skipped.")
} else {
  has_marker_gene_exp <- TRUE
}
```

@@

# create a data frame with raw exp and classification 
# will add this to the larger data frame
marker_gene_class_df <- markers_df |> 
  dplyr::select(barcodes, sum_raw_exp) |> 
  dplyr::distinct() |> 
  dplyr::mutate(
    marker_gene_classification = dplyr::if_else(
      barcodes %in% tumor_cells,
      "Tumor",
      "Normal"

    )
  )

# add classification to main classification data frame
classification_df <- classification_df |> 
  dplyr::left_join(marker_gene_class_df, by = "barcodes") |> 
  tidyr::pivot_longer(ends_with("classification"),
                      names_to = "method",
                      values_to = "classification") |> 
  dplyr::mutate(
    # make a factor so tumor always appears first 
    classification = forcats::fct_relevel(classification, "Tumor")
  )
```

@@

```{r, eval=has_marker_gene_exp}
# create a density plot showing the distribution of marker gene expression across classification methods 
classification_df |> 
  ggplot(aes(x = sum_raw_exp, color = classification)) +
  geom_density() +
  facet_wrap(vars(method),
             scales = "free_y") +
  scale_color_manual(values = plot_colors) +
  labs(
    x = "Raw sum of marker gene logcounts",
    color = ""
  ) 
```

@@

```{r}
# create matrix with marker genes as rows and barcodes as columns
marker_gene_matrix <- markers_df |> 
  dplyr::select(gene_expression, gene_symbol, barcodes) |> 
  tidyr::pivot_wider(values_from = gene_expression, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("gene_symbol") |> 
  as.matrix()

annotation_df <- classification_df |> 
  dplyr::select(barcodes, classification, method) |> 
  tidyr::pivot_wider(names_from = method, 
                     values_from = classification)

annotation <- ComplexHeatmap::columnAnnotation(
  AUCell = annotation_df$auc_classification, 
  marker_genes = annotation_df$marker_gene_classification,
  col = list(AUCell = plot_colors,
             marker_genes = plot_colors)
)
```

```{r, eval=has_marker_gene_exp}
# plot heatmap of marker genes 
plot_gene_heatmap(marker_gene_matrix, 
                  row_title = "Marker gene symbol",
                  legend_title = "Marker gene \nexpression",
                  annotation = annotation)
```

@@

# make density plot showing distribution of gene set scores for tumor/normal 
ggplot(geneset_plot_df, aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(rows = vars(geneset),
             cols = vars(method),
             scales = "free_y") +
  scale_color_manual(values = plot_colors) +
  labs(
    x = "Gene set score",
    color = ""
  )
```

@@

```{r}
# make a matrix of gene set by barcode 
geneset_matrix <- geneset_plot_df |> 
  dplyr::select(mean_score, geneset, barcodes) |> 
  unique() |> 
  tidyr::pivot_wider(values_from = mean_score, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("geneset") |> 
  as.matrix()

# plot heatmap of gene set score 
plot_gene_heatmap(geneset_matrix,
                  annotation = annotation,
                  legend_title = "Gene set \nscore")
```

>>>>

# test if _any_ marker genes have expression
if (sum(markers_df$gene_expression) == 0) {
  has_marker_gene_exp <- FALSE
  message("No marker gene expression detected, so all plots of marker genes will be skipped.")
} else {
  has_marker_gene_exp <- TRUE
}
```

@@

# create a data frame with raw exp and classification
# will add this to the larger data frame
marker_gene_class_df <- markers_df |>
  dplyr::select(barcodes, sum_raw_exp) |>
  dplyr::distinct() |>
  dplyr::mutate(
    marker_gene_classification = dplyr::if_else(
      barcodes %in% tumor_cells,
      "Tumor",
      "Normal"
    )
  )

# add classification to main classification data frame
classification_df <- classification_df |>
  dplyr::left_join(marker_gene_class_df, by = "barcodes") |>
  tidyr::pivot_longer(ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |>
  dplyr::mutate(
    # make a factor so tumor always appears first
    classification = forcats::fct_relevel(classification, "Tumor")
  )
```

@@

```{r, eval=has_marker_gene_exp}
# create a density plot showing the distribution of marker gene expression across classification methods
classification_df |>
  ggplot(aes(x = sum_raw_exp, color = classification)) +
  geom_density() +
  facet_wrap(vars(method),
    scales = "free_y"
  ) +
  scale_color_manual(values = plot_colors) +
  labs(
    x = "Raw sum of marker gene logcounts",
    color = ""
  )
```

@@

```{r}
# create matrix with marker genes as rows and barcodes as columns
marker_gene_matrix <- markers_df |>
  dplyr::select(gene_expression, gene_symbol, barcodes) |>
  tidyr::pivot_wider(
    values_from = gene_expression,
    names_from = barcodes
  ) |>
  tibble::column_to_rownames("gene_symbol") |>
  as.matrix()

annotation_df <- classification_df |>
  dplyr::select(barcodes, classification, method) |>
  tidyr::pivot_wider(
    names_from = method,
    values_from = classification
  )

annotation <- ComplexHeatmap::columnAnnotation(
  AUCell = annotation_df$auc_classification,
  marker_genes = annotation_df$marker_gene_classification,
  col = list(
    AUCell = plot_colors,
    marker_genes = plot_colors
  )
)
```

```{r, eval=has_marker_gene_exp}
# plot heatmap of marker genes
plot_gene_heatmap(marker_gene_matrix,
  row_title = "Marker gene symbol",
  legend_title = "Marker gene \nexpression",
  annotation = annotation
)
```

@@

# make density plot showing distribution of gene set scores for tumor/normal
ggplot(geneset_plot_df, aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(
    rows = vars(geneset),
    cols = vars(method),
    scales = "free_y"
  ) +
  scale_color_manual(values = plot_colors) +
  labs(
    x = "Gene set score",
    color = ""
  )
```

@@

```{r}
# make a matrix of gene set by barcode
geneset_matrix <- geneset_plot_df |>
  dplyr::select(mean_score, geneset, barcodes) |>
  unique() |>
  tidyr::pivot_wider(
    values_from = mean_score,
    names_from = barcodes
  ) |>
  tibble::column_to_rownames("geneset") |>
  as.matrix()

# plot heatmap of gene set score
plot_gene_heatmap(geneset_matrix,
  annotation = annotation,
  legend_title = "Gene set \nscore"
)
```
